Indirect elicitation as a necessary condition for consistent surrogate losses

### 2020-06-03 neurips-20/cvx-flats.tex
Meeting edits

\begin{definition}[Property, elicits]
	A \emph{property} is a set-valued function $\Gamma : \simplex \to 2^\R \setminus \emptyset$, which we denote $\Gamma: \simplex \toto \R$.
	A loss $L : \R \times \Y \to \reals$ \emph{elicits} the property $\Gamma$ if
	\begin{equation}
    \label{eq:elic}    
    \forall p \in \simplex, \;\; \Gamma(p) = \argmin_{u \in \R} \exploss{L}{u}{p}~.
	\end{equation}
\end{definition}

@@

An important caveat to the above definitions is that we have required $\Gamma = \prop{L}$ to be nonempty everywhere, meaning that $\exploss{L}{\cdot}{p}$ always achieves a minimum.
\raft{Let's make sure we agree with this paragraph}
While this condition is restrictive, and rules out several popular surrogates such as logistic and exponential loss for binary classification, these are easily rectified for example by thresholding the loss $L'(u,y) = \max(L(u,y),\epsilon)$ for sufficiently small $\epsilon>0$.
The results in this paper are not heavily impacted by this restriction, as for example assuming $\Gamma(p) \neq \emptyset$ whenever $p\in\relint\simplex$ is strong enough to include the above examples without impacting Theorems~\ref{thm:cvx-flats} and~\ref{thm:bayes-risk-lower-bound}.
\raft{In particular, I'm torn about whether this statement is accurate; in the sense that we can restrict to $\P\subseteq\simplex$ I think it's fair, but in the sense that calibration does not imply indirect elicitation without it, perhaps not.}
We expect that a generalization of property elicitation which allows for ``infinite'' predictions (e.g., along a prescribed ray), thereby ensuring a minimum is always achieved for convex losses, would allow us to lift this restriction entirely.
\btw{Some refs / related work commented out here.  I think not as relevant as elicitation complexity.}

@@

\begin{theorem}\label{thm:consistent-implies-indir-elic}
	If a surrogate and link pair $(L, \psi)$ is consistent with respect to a property $\gamma$ or loss $\ell$ eliciting $\gamma$, then $(L, \psi)$ indirectly elicits $\gamma$.
\end{theorem}
\begin{proof}
%  \raft{Maybe ref the consistency definition here explicitly, at the top and the end}
	By Lemma~\ref{lem:consistent-loss-implies-prop}, it suffices to show the result for consistency with respect to a property $\gamma$, setting $\gamma := \prop{\ell}$ if $\ell$ is given instead.
	We show the contrapositive; suppose $(L, \psi)$ does not indirectly elicit $\gamma$, meaning we have some $p \in \simplex$ so that $u \in \Gamma(p)$ but $\psi(u) \not \in \gamma(p)$, where $\Gamma := \prop{L}$.
	Observe that we use the fact $\Gamma(p) \neq \emptyset$.
	Consider the constant sequence $\{f_m\}$ with $f_m(x) = u$ for all $m,x$, and take $D$ with full support on some $x \in \X$, and let $D_x = p$.
  Since $u \in \Gamma(p)$, we observe $\E_D L(f_m(X), Y) = \inf_f L(f(X),Y)$ for all $m$; in particular $\E_D L(f_m(X), Y) \to \inf_f L(f(X),Y)$.
%  \raft{Maybe clearer to first say $\E_D L(f_m(X), Y) = \inf_f L(f(X),Y)$ for all $m$, and then the easy implication of the convergence \jessie{Done}}
	However, we have $\E_X \propdis(\psi \circ f_m(X), D_X) = \propdis(\psi(u_m), p) = \propdis(\psi(u), p) \neq 0$, since $\psi(u) \not \in \gamma(p)$.
%  \raft{Almost -- just clarify that this is the expectation over $D$ too by construction \jessie{Kinda done?}}
	Thus, we observe $(L, \psi)$ is not consistent with respect to $\gamma$ (Definition~\ref{def:consistent-prop}).
\end{proof}

@@

A hyperplane weakly separates two sets if its two closed halfspaces respectively contain the two sets.
\begin{lemma}\label{lem:intersect-levelsets}
  If $\gamma: \simplex \toto \R$ is an elicitable property, then for any pair of predictions $r \neq r' \in \R$, there is a hyperplane $H = \{p : v \cdot p = 0\}$ that weakly separates both $\gamma_r$ and $\gamma_{r'}$ with $\gamma_r \cap \gamma_{r'} \subseteq H$.
\end{lemma}
\begin{proof}
  Because $\gamma$ is elicitable, its level sets are closed convex bounded sets.\bo{cite}
  Note that if the intersection $\gamma_r \cap \gamma_{r'} = \emptyset$, then the separating hyperplane theorem already gives an $H$ that strictly separates the two sets, in particular, $\gamma_r \cap \gamma_{r'}$.

  \bo{Not finished - Sketch next. I don't know if we can appeal to something, but here's my approach to proceed.}

  \bo{Approach 1}

  Each level set corresponds to a report $r$, which corresponds to a supporting hyperplane $h_r$ of the epigraph of the Bayes risk, where $\gamma_r$ is the set of points above which $h_r$ supports $\epi(\ellbar)$.\bo{cite...?}
  So $\gamma_r \cap \gamma_{r'}$ is the set of points at which both $h_r$ and $h_{r'}$ support $\epi(\ellbar)$.
  \bo{Idea: use properties of hyperplanes from here...not sure how exactly}

  \bo{Approach 2}

  First, we claim $\gamma_r$ and $\gamma_{r'}$ are either equal, or have disjoint relative interiors. \bo{True?}

  If not equal, then the intersection is on the relative boundary of both of them, which since they're convex should give a weakly separating hyperplane.
\end{proof}

\begin{lemma}\label{lem:set-valued-prop-flats}
	Suppose we are given an elicitable property $\gamma : \simplex \toto \R$ and distribution $p \in \relint\simplex$ such that $p \in \gamma_r \cap \gamma_{r'}$ for $r,r' \in \R$.
	Then for any flat $F$ containing $p$, $F \cap \simplex \subseteq \gamma_r \iff F \cap \simplex \subseteq \gamma_{r'}$.
\end{lemma}
\begin{proof}
  First, by Lemma \ref{lem:intersect-levelsets}, $\gamma_r \cap \gamma_{r'}$ is a closed convex subset of a hyperplane that weakly separates $\gamma_r$ and $\gamma_{r'}$.
  Let it be parameterized by the unit vector $v \in \reals^{\Y}$, where $\gamma_r \subseteq \{ q \in \simplex : v \cdot q \leq 0 \}$, while $\gamma_{r'} \subseteq \{ q \in \simplex : v \cdot q \geq 0\}$, and $q \in \gamma_r \cap \gamma_{r'} \implies v \cdot q = 0$.

  Now, we prove the equivalent claim $F \cap \simplex \subsetneq \gamma_r \iff F \cap \simplex \subsetneq \gamma_{r'}$.
  Suppose $F \cap \simplex$ has a point $q$ not in $\gamma_r$.
  If $v \cdot q < 0$, then $q$ is already not in $\gamma_{r'}$, QED.
  Otherwise, we claim that $p - \epsilon (q-p) \in F \cap \simplex$ for small enough $\epsilon$.
  Containment in $F$ follows because both $p$ and $q$ are in $F$ and it is a flat, while containment in $\simplex$ follows because $p \in \relint\simplex$.
  But $v \cdot (p - \epsilon (q-p)) = -\epsilon (v \cdot q) < 0$, so $q$ is not in $\gamma_{r'}$, QED.
\end{proof}

@@

\begin{theorem}\label{thm:cvx-flats}
  Let a property $\gamma:\simplex\toto\R$ be given, let $p\in\simplex$ such that either (i) $|\gamma(p)|=1$ or (ii) $\gamma$ is elicitable and $p\in\relint\simplex$, and let $r\in\gamma(p)$.
Then if $L: \reals^d \times \Y \to \reals$ indirectly elicits $\gamma$, there is a flat $F$ with $p\in F \cap \simplex \subseteq \gamma_r$ and $\codim(F) \leq d$.
% COmmented 06.03.20
%	If there is a convex surrogate loss $L : \reals^d \times\Y \to \reals$ indirectly eliciting $\gamma$, then for all $p \in \simplex$ and $r \in \R$ such that $r \in \gamma(p)$, there is a flat $F \subseteq \affhull(\simplex)$ of codimension at most $d$ such that $F \cap \simplex \subseteq \gamma_r$ and $p \in F$.
\end{theorem}
\begin{proof}
	Let $\Gamma := \prop{L}$ and suppose $(L, \psi)$ indirectly elicits $\gamma$, so that for all $u \in \reals^d$, we have $\Gamma_u \subseteq \gamma_{\psi(u)}$.
	We now claim there is some $u \in \Gamma(p)$ with $\psi(u) = r$.
	First, if $|\gamma(p)| = 1$, then any $u \in \Gamma(p)$ must link to $r$ as we contradict indirect elicitation otherwise; we would have $p \in \Gamma_u$ but $p \not \in \gamma_{\psi(u)}$.
	Now if $\gamma$ is elicitable and $p \in \relint{\simplex}$, we know that $\psi(u) = r'$ for some $r' \in \R$.%suppose there is another $r' \in \R$ with $r \neq r'$ and $r' \in \gamma(p)$.
	If $r' \neq r$, we can show $p \in F \cap \simplex \subseteq \gamma_{r'} (\iff p \in F \cap \simplex \subseteq \gamma_{r})$, where the if and only if comes from Lemma~\ref{lem:set-valued-prop-flats}.
%	Fix $p \in \simplex$ and $r \in \gamma(p)$.
  \jessie{Show that there is some $u \in \Gamma(p)$ such that $\psi(u) = r$.}
	We will show $F \cap \simplex \subseteq \Gamma_u$; the result then follows from $\Gamma_u \subseteq \gamma_r$.
	As $L$ is convex and elicits $\Gamma$, we have $u \in \Gamma(p) \iff \vec 0 \in \partial \exploss{L}{u}{p} \iff \vec 0 \in \oplus_y p_y \partial L(u,y)$, where $\oplus$ denotes the Minkowski sum~\citep[Theorem 4.1.1]{hiriart2012fundamentals}.
  % (Observe that in order for $L$ to be convex, $L(\cdot, y)$ must be convex for all $y \in \Y$.)

>>>>

\begin{definition}[Property, elicits]
	A \emph{property} is a set-valued function $\Gamma : \simplex \to 2^\R \setminus \{\emptyset\}$, which we denote $\Gamma: \simplex \toto \R$.
	A loss $L : \R \times \Y \to \reals$ \emph{elicits} the property $\Gamma$ if
	\begin{equation}
    \label{eq:elic}    
    \forall p \in \simplex, \;\; \Gamma(p) = \argmin_{u \in \R} \exploss{L}{u}{p}~.
	\end{equation}
\end{definition}

@@

An important caveat to the above definitions is that we have required $\Gamma = \prop{L}$ to be nonempty everywhere, meaning that $\exploss{L}{\cdot}{p}$ always achieves a minimum.
\raft{Let's make sure we agree with this paragraph}
While this condition technically rules out several popular surrogates such as logistic and exponential loss for binary classification, these can be included for example by thresholding the loss $L'(u,y) = \max(L(u,y),\epsilon)$ for sufficiently small $\epsilon>0$.
\botodo{work on ``heavily impacted''}
The results in this paper are not heavily impacted by this restriction, as for example assuming $\Gamma(p) \neq \emptyset$ whenever $p\in\relint\simplex$ is strong enough to include the above examples without impacting Theorems~\ref{thm:cvx-flats} and~\ref{thm:bayes-risk-lower-bound}.
\raft{In particular, I'm torn about whether this statement is accurate; in the sense that we can restrict to $\P\subseteq\simplex$ I think it's fair, but in the sense that calibration does not imply indirect elicitation without it, perhaps not.}
We expect that a generalization of property elicitation which allows for ``infinite'' predictions (e.g., along a prescribed ray), thereby ensuring a minimum is always achieved for convex losses, would allow us to lift this restriction entirely.
\btw{Some refs / related work commented out here.  I think not as relevant as elicitation complexity.}

@@

\begin{theorem}\label{thm:consistent-implies-indir-elic}
	\botodo{Assume L is minnable}If a surrogate and link pair $(L, \psi)$ is consistent with respect to a property $\gamma$ or loss $\ell$ eliciting $\gamma$, then $(L, \psi)$ indirectly elicits $\gamma$.
\end{theorem}
\begin{proof}
%  \raft{Maybe ref the consistency definition here explicitly, at the top and the end}
	By Lemma~\ref{lem:consistent-loss-implies-prop}, it suffices to show the result for consistency with respect to a property $\gamma$, setting $\gamma := \prop{\ell}$ if $\ell$ is given instead.
	We show the contrapositive; suppose $(L, \psi)$ does not indirectly elicit $\gamma$, meaning we have some $p \in \simplex$ so that $u \in \Gamma(p)$ but $\psi(u) \not \in \gamma(p)$, where $\Gamma := \prop{L}$.
	Observe that we use the fact $\Gamma(p) \neq \emptyset$.
	Consider the constant sequence $\{f_m\}$ with $f_m(x) = u$ for all $m,x$, and take $D$ with full support on some $x \in \X$, and let $D_x = p$.
  Since $u \in \Gamma(p)$, we observe $\E_D L(f_m(X), Y) = \inf_f L(f(X),Y)$ for all $m$; in particular $\E_D L(f_m(X), Y) \to \inf_f L(f(X),Y)$.
%  \raft{Maybe clearer to first say $\E_D L(f_m(X), Y) = \inf_f L(f(X),Y)$ for all $m$, and then the easy implication of the convergence \jessie{Done}}
	However, we have $\E_X \propdis(\psi \circ f_m(X), D_X) = \propdis(\psi(u_m), p) = \propdis(\psi(u), p) \neq 0$, since $\psi(u) \not \in \gamma(p)$.
%  \raft{Almost -- just clarify that this is the expectation over $D$ too by construction \jessie{Kinda done?}}
	Thus, we observe $(L, \psi)$ is not consistent with respect to $\gamma$ (Definition~\ref{def:consistent-prop}).
\end{proof}

@@

A hyperplane weakly separates two sets if its two closed halfspaces respectively contain the two sets.
\begin{lemma}\label{lem:intersect-levelsets}
  If $\gamma: \simplex \toto \R$ is an elicitable property, then for any pair of predictions $r, r' \in \R$ where $\gamma_r \neq \gamma_{r'}$, there is a hyperplane $H = \{x \in \reals^{\Y} : v \cdot x = 0\}$, for some $v \in \reals^\Y$, that weakly separates both $\gamma_r$ and $\gamma_{r'}$ with $\gamma_r \cap H = \gamma_{r'} \cap H = \gamma_r \cap \gamma_{r'}$.
\end{lemma}
\begin{proof}
  Because $\gamma$ is elicitable, its level sets are closed convex bounded sets.\bo{cite}
  Note that if the intersection $\gamma_r \cap \gamma_{r'} = \emptyset$, then the separating hyperplane theorem already gives an $H$ that strictly separates the two sets, in particular, $\gamma_r \cap \gamma_{r'}$.

  Suppose $\gamma$ is elicited by $\ell$.
  Let $v = \ell(r, \cdot) - \ell(r', \cdot)$, interpreted as a vector in $\reals^\Y$.
  Let $H = \{ q : v \cdot q = 0 \}$.
  If $v \cdot q < 0$, then $r'$ cannot be optimal, so $q \not\in \gamma_{r'}$.
  So $\gamma_{r'} \subseteq \{ q : v \cdot q \geq 0 \}$.
  Symmetrically, $\gamma_r \subseteq \{ q : v \cdot q \leq 0 \}$.

  Finally, if and only if $v \cdot q = 0$, i.e. $q \in H$, by definition the expected losses of both reports are the same.
  So $q \in \gamma_r \cap H \iff q \in \gamma_{r'} \cap H$.
\end{proof}

\begin{lemma}\label{lem:set-valued-prop-flats}
	Suppose we are given an elicitable property $\gamma : \simplex \toto \R$ and distribution $p \in \relint\simplex$ such that $p \in \gamma_r \cap \gamma_{r'}$ for $r,r' \in \R$.
	Then for any flat $F$ containing $p$, $F \cap \simplex \subseteq \gamma_r \iff F \cap \simplex \subseteq \gamma_{r'}$.
\end{lemma}
\begin{proof}
  If $\gamma_r = \gamma_{r'}$, we are done.
  Otherwise:
  First, by Lemma \ref{lem:intersect-levelsets}, $\gamma_r \cap \gamma_{r'}$ is a closed convex subset of a hyperplane that weakly separates $\gamma_r$ and $\gamma_{r'}$.
  Let it be parameterized by the unit vector $v \in \reals^{\Y}$, where $\gamma_r \subseteq \{ q \in \simplex : v \cdot q \leq 0 \}$, while $\gamma_{r'} \subseteq \{ q \in \simplex : v \cdot q \geq 0\}$, and $q \in \gamma_r \cap \gamma_{r'} \implies v \cdot q = 0$.

  \bo{Also get that only points in hyperplane are in the intersection}
  Now take a point $q \in \gamma_r \setminus \gamma_{r'}$ (WLOG, swapping roles if necessay).
  By containment in $\gamma_r$, we have $v \cdot q \leq 0$.
  Then $v \cdot q < 0$, because if $q$ is in the hyperplane and in $\gamma_r$, then it is in $\gamma_{r'}$.
  Now we claim that $p - \epsilon (q-p) \in F \cap \simplex$ for small enough $\epsilon$.
  Containment in $F$ follows because both $p$ and $q$ are in $F$ and it is a flat, while containment in $\simplex$ follows because $p \in \relint\simplex$.
  But $v \cdot (p - \epsilon (q-p)) = -\epsilon (v \cdot q) < 0$, so $q$ is not in $\gamma_{r'}$, QED.
\end{proof}

@@

\begin{theorem}\label{thm:cvx-flats}
  Let a property $\gamma:\simplex\toto\R$ be given, let $p\in\simplex$ such that either (i) $|\gamma(p)|=1$ or (ii) $\gamma$ is elicitable and $p\in\relint\simplex$, and let $r\in\gamma(p)$.
Then if $L: \reals^d \times \Y \to \reals$ indirectly elicits $\gamma$, there is a flat $F$ with $p\in F \cap \simplex \subseteq \gamma_r$ and $\codim(F) \leq d$.
% COmmented 06.03.20
%	If there is a convex surrogate loss $L : \reals^d \times\Y \to \reals$ indirectly eliciting $\gamma$, then for all $p \in \simplex$ and $r \in \R$ such that $r \in \gamma(p)$, there is a flat $F \subseteq \affhull(\simplex)$ of codimension at most $d$ such that $F \cap \simplex \subseteq \gamma_r$ and $p \in F$.
\end{theorem}
\begin{proof}

%	There is some u in Gamma(p). (Gamma nonempty.)
%	Gamma_u contains a big ol flat F.
%
%	Let r' = psi(u).
%	Hence gamma_{r'} contains F.
%	If singleton, we know r' = r, so done.
%	Otherwise, elicitable and p in relint.
%	If r' = r, done. Otherwise:
%	We can apply Lemma 3: p is in gamma_r cap gamma_{r'}, F contains p and is in gamma_{r'}.
%	So F is also in gamma_r.

	Let $\Gamma := \prop{L}$ and suppose $(L, \psi)$ indirectly elicits $\gamma$, so that for all $u \in \reals^d$, we have $\Gamma_u \subseteq \gamma_{\psi(u)}$.
	We now claim there is some $u \in \Gamma(p)$ with $\psi(u) = r$.
	First, if $|\gamma(p)| = 1$, then any $u \in \Gamma(p)$ must link to $r$ as we contradict indirect elicitation otherwise; we would have $p \in \Gamma_u$ but $p \not \in \gamma_{\psi(u)}$.
	Now if $\gamma$ is elicitable and $p \in \relint{\simplex}$, we know that $\psi(u) = r'$ for some $r' \in \R$.%suppose there is another $r' \in \R$ with $r \neq r'$ and $r' \in \gamma(p)$.
	If $r' \neq r$, we can show $p \in F \cap \simplex \subseteq \gamma_{r'} (\iff p \in F \cap \simplex \subseteq \gamma_{r})$, where the if and only if comes from Lemma~\ref{lem:set-valued-prop-flats}.
%	Fix $p \in \simplex$ and $r \in \gamma(p)$.
  \jessie{Show that there is some $u \in \Gamma(p)$ such that $\psi(u) = r$.}
	We will show $F \cap \simplex \subseteq \Gamma_u$; the result then follows from $\Gamma_u \subseteq \gamma_r$.
	As $L$ is convex and elicits $\Gamma$, we have $u \in \Gamma(p) \iff \vec 0 \in \partial \exploss{L}{u}{p} \iff \vec 0 \in \oplus_y p_y \partial L(u,y)$, where $\oplus$ denotes the Minkowski sum~\citep[Theorem 4.1.1]{hiriart2012fundamentals}.
  % (Observe that in order for $L$ to be convex, $L(\cdot, y)$ must be convex for all $y \in \Y$.)
